The ``Dark Side'' of Phylogenetic Comparative Methods

### 2015-10-09 Manuscript/dark.tex
Play with solutions

Point 1: set the context and purpose for the work; 
Point 2: indicate the approach and methods used; 
Point 3: outline the main results; 
Point 4: identify the conclusions, the wider implications and the relevance to management or policy.

@@

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any other statistical method, and this is well established in the literature \citep[e.g.][]{freckleton2009seven,boettiger2012your,losos2011seeing,blomberg2012independent}.
Increasingly however, assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results (see examples below). % ADD EXAMPLES
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing,westoby1995misinterpreting}.

@@

\subsection{Examples of the problem}
\subsubsection{Trait-dependent diversification}

@@

\subsubsection{Phylogenetically independent contrasts}
Even the most commonly used phylogenetic comparative method, phylogentically independent contrasts has weel-known assumptions that are rarely tested.

@@

\subsection{How can we solve some of these problems?}

\subsubsection{1. Simplify, summarize and share}
Many of the problems above are caused by methods papers being difficult to read. 
This isn't entirely the fault of methods-developers; most journals have strict word limits so it's often easier to use a one line equation rather than a paragraph of description. 
However, it's hard to complain about novice users failing to understand a method when the paper is incomprehensible. %cough cough Hansen. Ow my eyes are bleeding.
One solution is to prepare an accompanying blog post or video explaining the method in less technical terms. 
Some journals already encourage this (e.g. Methods in Ecology and Evolution), and even if they do not, there is no harm in doing this and hosting it on a personal website. 
The ability to share ideas with a non-technical audience is a key skill to develop, and it may even encourage more people to use the method. 
% NC: Perhaps add something about having reproducible knitr reports in supp mat (e.g. cite that nice recent paper)
At the very least, having a simple list of the main assumptions, biases and caveats of the method somewhere would reduce misuse and provide a place to point methods-users to when they begin using a method.
Of course this will be difficult as there are often hidden assumptions in a method, and listing all possible assumptions and caveats may be unfeasible.\\

The glut of literature that already exists for established methods is a bigger problem. 
One solution would be to establish a wiki-style website where methods-developers and advanced methods-users could post summaries of commonly used methods, along with lists of key papers to read. 
The community would be responsible for peer reviewing these summaries to ensure all opinions are covered. 
Erick Matsen's Phylobabble discourse page (http://phylobabble.org/) is an excellent example of a similar approach for phylogenetics. Unfortunately there is currently little incentive for methods-developers and advanced methods-users to contribute to such a project at the cost of time pursuing their own research.\\

One solution is to establish a discourse page for providing R help. 
Currently most methods-users either email the R-sig-phylo help list or the package developer for assistance. 
Both approaches are suboptimal. 
R-sig-phylo is impossible to search effectively, leading to the same questions being dealt with repeatedly. 
The same is true in email exchanges between methods-users and package developers with the added issue of this being ``dark advice'' that isn't shared. 
A discourse page would collate all of this information and reduce pressure on package developers by allowing advanced methods-users to help novice users with simple queries. 
We are in the process of setting up a discourse page for this purpose (it will be called ``phyloBootstrap'' because it will provide ``support'' for PCMs). At Evolution 2014 many methods-developers agreed to move their R package help to this forum.

\subsubsection{2. Read, think and learn}
If methods-developers are going to put time and effort into making their methods easier to understand, then methods-users need to take advantage of this and work harder to understand the methods they are using. 
At a minimum, methods-users should read the original papers describing a method, and any recent updates. 
Jumping straight into an R implementation is a tolerable way to start a project, but it needs to be followed up by careful reading and thought. 
Methods-users also need to avoid retrofitting their question to the newest methods, instead they should think carefully about the question they want to ask, whether the method is appropriate for the question, and whether PCMs are
needed at all \citep{losos2011seeing}.\\

Another key skill to develop is distrust; never take results from PCMs (or any other statistical analysis) at face value. 
A good way to check a method is to simulate some data and see if the results are as expected. 
This can expose hidden assumptions or biases that have not been explored in the papers accompanying the method, or reveal a lack of understanding of the the mechanics of the method being used. 
It is also important for methods-users to consider whether the method will work on their dataset; for example, how many species does it need to have reasonable power, how does it deal with polytomies, can it cope with non-ultrametric trees etc.
Finally, methods-users should never be afraid to question standard practice, sometimes it's just PCM folklore!
% NC: Could add my pglm vs pgls example.

\subsubsection{3. Collaboration}
A obvious solution to all these problems is for methods-developers and methods-users to collaborate more. 
The benefits to methods-users are obvious, but methods-developers can also benefit by gaining extra datasets to test their ideas on and people who will discover corner cases and bugs in their software before it get released. 
A big difficulty with collaboration is that it is often asymmetrical in terms of benefit to methods-users and methods-developers. 
Methods-developers and advanced methods-users can often put a lot of time and effort into running analyses for less experienced colleagues, yet at best they come out of the experience with a middle authored paper. 
Until these author positions are considered equal to first and last author positions, there is little incentive to collaborate extensively.

\subsubsection{4. Incentives}
It is important to recognize that our ability to do rigorous quantitative science often relies on highly-skilled methods-developers, especially as evolutionary biology becomes ever more computationally intensive. 
We cannot afford to lose these people to industry, nor can we afford to pay industry wages, thus we need to make it worthwhile for methods-developers to remain in (or at least interact with) academia.

First we need to stop insisting that methods papers must be entirely novel. 
Improvements to existing methods, and ways of detecting biases or testing model fit should be sufficient for publication. 
This is fairly standard in other fields, e.g. statistical phylogenetics, and these kinds of papers are arguably more useful to the community than constantly publishing new methods.
When novel methods are published, journals should encourage methods-developers to include ways of testing the assumptions of their methods within the original publications and packages, and request simplified summaries to accompany technical papers.

Second we need to fund pure methods development, including incremental methods. 
Currently it is difficult to get funding for purely methods driven research; an empirical component is always needed and methods development is often seen as part of the bigger empirical picture, rather than the reverse. 
This unfairly prejudices funding bodies (and hiring committees) away from methods-developers.

Finally we need to re-evaluate our criteria for success in academia. 
The ``publish or perish'' mentality does not encourage methods-developers to spend more time on their methods than necessary. 
Thus there is little incentive to summarize, simplify and share, or to collaborate with methods-users, as we suggest above. 
This mentality is also not conducive to methods-users attempting to fully understand a method prior to using it. 
Instead we need to value quality over quantity when it comes to publications.
This would improve career prospects for both methods-developers and careful methods-users.

\section{Conclusion}
% NC: Still need a couple of lines of conclusion but I'm drawing a blank. Final sentence is fine I think. We could just skip the conclusion heading and have this as a standalone sentence at the end.

\section{Acknowledgments}

Thanks to the Society of Systematic Biologists for funding the symposium at Evolution 2014, in Raleigh, NC, where the ideas in this paper were first consolidated. This work was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696 (NC).

\section{Data Accessibility}
All data are available on GitHub at https://github.com/nhcooper123/pcm-darkside.

>>>>

Phylogenetic comparative methods are becoming increasingly popular for investigating evolutionary patterns and processes.
However, these methods are not infallible - they suffer from biases and make assumptions like all other statistical methods.
Unfortunately, although these limitations are often well-known in the phylogenetic comparative methods community, they are often inadequately assessed in empirical studies leading to misinterpreted results and poor model fits.
Here we explore reasons for this communication gap between those developing new methods and those using them. 
We suggest that some important pieces of information are missing from the literature, and that others are difficult to extract from long, technical papers.
We also highlight problems with users jumping straight into implementations (e.g. in R) that may lack documentation of biases and assumptions that are mentioned in the original papers.
To help solve these problems we propose...

@@

Unfortunately, PCMs also have a ``dark side''; they make various assumptions and suffer from biases in exactly the same way as any other statistical method, and this is well established in the literature \citep[e.g.][]{freckleton2009seven,boettiger2012your,losos2011seeing,blomberg2012independent}.
Increasingly however, assumptions and biases are inadequately assessed in empirical studies, leading to poor model fits and misinterpreted results (see examples below).
Additionally, little consideration is given to whether using a PCM is really appropriate for the question at hand \citep{losos2011seeing,westoby1995misinterpreting}.

@@

\subsection{Examples of the problem}
\subsubsection{Trait-dependent diversification} % FINISH and get GAVIN TO CHECK!!!

@@

\subsubsection{Phylogenetically independent contrasts} % FINISH!!!
Even the most commonly used phylogenetic comparative method, phylogentically independent contrasts has weel-known assumptions that are rarely tested.

@@

\subsection{How can we solve some of these problems?}
  \subsubsection{1. Simplify, summarize and share}
    Many of the problems above are caused by methods papers being difficult to read. 
    This is not entirely the fault of those writing these papers; most journals have strict word limits so it is often easier to use a one line equation rather than a paragraph of description. 
    However, it is hard to complain about novice users failing to understand a method when the paper is incomprehensible.
    One solution is to prepare an accompanying blog post or video explaining the method in less technical terms. 
    Some journals already encourage this (including the British Ecological Society journals; see https://soundcloud.com/besjournals, https://www.youtube.com/user/MethodsEcolEvol), and even if they do not, there is no harm in doing this and hosting it on a personal website. 
    The ability to share ideas with a non-technical audience is a key skill to develop, and may encourage more people to use the method.
    Encouraging increased efforts at reproducibility would also help, by including knitR \citep{Xie:2015aa} reports in Supplementary Material showing exactly how each analysis was run \citep[e.g.][]{fitzjohn2014much}, or by requiring all analyses and code to be available on GitHub or Bitbucket. 
    At the very least, having a list of the main assumptions, biases and caveats of the method somewhere would reduce misuse and provide a place to point people to when they begin using a method (this will be difficult as there are often hidden assumptions in a method, and listing all possible assumptions and caveats may be unfeasible in some cases).\\

    The glut of literature that already exists for established methods is a bigger problem. 
    One solution would be to establish a wiki-style website where people could post summaries of commonly used methods, along with lists of key papers to read. 
    The community would be responsible for peer reviewing these summaries to ensure all opinions are covered. 
    Erick Matsen's Phylobabble discourse page (http://phylobabble.org/) is an excellent example of a similar approach for phylogenetics. 
    The British Ecological Society's Quantitative Ecology Special Interest Group is also in the process of creating a ``Field Guide for Ecologists'' (http://bes-qsig.github.io/fge/) that will fill a similar niche in ecological methods.
    Unfortunately there is currently little incentive for people to contribute to such a project at the cost of time pursuing their own research.\\

%One solution is to establish a discourse page for providing R help. 
%Currently most methods-users either email the R-sig-phylo help list or the package developer for assistance. 
%Both approaches are suboptimal. 
%R-sig-phylo is impossible to search effectively, leading to the same questions being dealt with repeatedly. 
%The same is true in email exchanges between methods-users and package developers with the added issue of this being ``dark advice'' that isn't shared. 
%A discourse page would collate all of this information and reduce pressure on package developers by allowing advanced methods-users to help novice users with simple queries. 
%We are in the process of setting up a discourse page for this purpose (it will be called ``phyloBootstrap'' because it will provide ``support'' for PCMs). At Evolution 2014 many methods-developers agreed to move their R package help to this forum.

  \subsubsection{2. Be more cynical about the methods you use}
    A key skill to develop in science is cynicism - never take results from PCMs (or any other statistical analysis) at face value.
    At a minimum, it is worth reading the original papers describing a method, plus any recent updates, and looking carefully for assumptions and caveats that may affect the analyses at hand.
    A good way to check a method is to simulate some data and see if the results are as expected \citep[e.g.][]{boettiger2012your}. 
    This can expose hidden assumptions or biases that have not been explored in the papers accompanying the method, or reveal a lack of understanding of the the mechanics of the method being used. 
    It is also important to consider whether the method will work on a particular dataset for example.
    \begin{itemize}
      \item How many species does it need to have reasonable power? 
      Often methods require more species than are generally available. 
      For example, the new trait-dependent diversification method of \citet{rabosky2015robust} is ``primarily be applicable to phylogenies that include at least several thousand tips'' although the authors suggest that most such analyses have \textless 1000 tips.
      \item How does it deal with polytomies?
      Many methods arbitrarily resolve polytomies using zero length branches, thus polytomies can inflate rates of evolution, and bias models of evolution \citep{cooper2010body}.
      \item Can the method use non-ultrametric trees? 
      Not all PCMs can use non-ultrametric trees. 
      For example, some current implementations of the OU model should not be used with non-ultrametric trees because they are based on transforming the tree directly, rather than transforming the variance co-variance matrix (e.g. MOTMOT; \citealp{Thomas:2011aa}), and non-ultrametric trees cause variance-covariance matrices to become non-tree like \citep{slater2014correction}. 
    \end{itemize}
    It is also important to avoid retrofitting questions to the newest methods, instead think carefully about the question, whether the method is appropriate for the question, and whether PCMs are needed at all \citep{losos2011seeing}.
    Finally, methods-users should never be afraid to question standard practice, sometimes it's just PCM folklore!

%\subsubsection{3. Collaboration}
%A obvious solution to all these problems is for methods-developers and methods-users to collaborate more. 
%The benefits to methods-users are obvious, but methods-developers can also benefit by gaining extra datasets to test their ideas on and people who will discover corner cases and bugs in their software before it get released. 
%A big difficulty with collaboration is that it is often asymmetrical in terms of benefit to methods-users and methods-developers. 
%Methods-developers and advanced methods-users can often put a lot of time and effort into running analyses for less experienced colleagues, yet at best they come out of the experience with a middle authored paper. 
%Until these author positions are considered equal to first and last author positions, there is little incentive to collaborate extensively.

  \subsubsection{3. Incentives}
    It is important to recognize that our ability to do rigorous quantitative science often relies on highly-skilled methods-developers, especially as evolutionary biology becomes ever more computationally intensive. 
    We cannot afford to lose these people to industry, nor can we afford to pay industry wages, thus we need to make it worthwhile for such skilled researchers to remain in (or at least interact with) academia.

    First we need to stop insisting that methods papers are entirely novel. 
    Improvements to existing methods, and ways of detecting biases or testing model fit should be sufficient for publication.
    This is fairly standard in other fields, e.g. statistical phylogenetics, and these kinds of papers are arguably more useful to the community than constantly publishing new methods.
    When novel methods are published, journals should encourage researchers to include ways of testing the assumptions of their methods within the original publications and packages, and request simplified summaries to accompany technical papers.

    Second we need to fund pure methods development, including incremental methods. 
    Currently it is difficult to get funding for purely methods driven research; an empirical component is always needed and methods development is often seen as part of the bigger empirical picture, rather than the reverse. 
    This unfairly prejudices funding bodies (and hiring committees) away from this kind of research.

    Finally, an obvious solution to many of these problems is for methods-developers and methods-users to collaborate more. 
    The benefits to methods-users are obvious, but methods-developers can also benefit by gaining extra datasets to test their ideas on and people who will discover corner cases and bugs in their software before it get released. 
    A big difficulty with collaboration is that it is often asymmetrical in terms of benefit; some researchers can put a lot of time and effort into running analyses for less experienced colleagues, yet at best they come out of the experience with a middle authored paper. 
    Until these author positions are considered equal to first and last author positions, there is little incentive to collaborate extensively.

  \subsection{Conclusion}

%I guess I’d close by saying it’s our responsbilitiy as part of the methods developing and using community to find innovative ways to tackle these challenges.

\subsection{Acknowledgments}
This paper is based on a talk given by NC at the Methods in Ecology and Evolution 5th Anniversary Symposium.
Thanks to the Society of Systematic Biologists for funding the symposium at Evolution 2014 where the ideas in this paper were first consolidated. 
This work was supported by The European Commission CORDIS Seventh Framework Program (FP7) Marie Curie CIG grant, proposal number: 321696 (NC).

\subsection{Data Accessibility}
All data are available on GitHub at https://github.com/nhcooper123/pcm-darkside.
